---
title: "Homework 05"
engine: knitr
format:
  pdf:
    pdf-engine: xelatex
    papersize: letter
    fontsize: 10pt
    geometry: margin=0.75in
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
        \RecustomVerbatimEnvironment{verbatim}{Verbatim}{breaklines}
        \usepackage{etoolbox}
        \BeforeBeginEnvironment{verbatim}{\par\noindent\begin{minipage}{\linewidth}}
        \AfterEndEnvironment{verbatim}{\end{minipage}\par}
execute:
  warning: false
  message: false
---

## Question 1: Estimate the bias and variance of KNN

### (a)

The true mean response at the target is

$$
f(\mathbf{x}_0) = 0.5(0.5) + \sin(0.7) - 0.3(1)^2
\approx 0.594218.
$$

```{r}
library(FNN)
set.seed(43251)
n = 400
p = 3
k_values = seq(1, 29, 2)
x0 = matrix(c(0.5, 0.7, 1), nrow = 1)
f0 = 0.5*x0[1,1] + sin(x0[1,2]) - 0.3*x0[1,3]^2
X = matrix(rnorm(n*p), nrow = n)
y = 0.5*X[,1] + sin(X[,2]) - 0.3*X[,3]^2 +
  rnorm(n, sd = 0.1)
pred = numeric(length(k_values))
for (m in seq_along(k_values)) {
  knn.fit = knn.reg(X, x0, y, k = k_values[m], algorithm = "brute")
  pred[m] = knn.fit$pred
}
data.frame(k = k_values, prediction = pred, truth = f0)
```

### (b)

Each of the 200 repetitions uses one new training dataset for all values
of $k$. Rows of `pred_matrix` represent repetitions; columns represent $k$.

```{r}
set.seed(43252)
nsim = 200
pred_matrix = matrix(NA_real_, nrow = nsim, ncol = length(k_values))
for (l in seq_len(nsim)) {
  X = matrix(rnorm(n*p), nrow = n)
  y = 0.5*X[,1] + sin(X[,2]) - 0.3*X[,3]^2 +
    rnorm(n, sd = 0.1)
  for (m in seq_along(k_values)) {
    knn.fit = knn.reg(X, x0, y, k = k_values[m], algorithm = "brute")
    pred_matrix[l, m] = knn.fit$pred
  }
}
mean_pred = colMeans(pred_matrix)
bias = mean_pred - f0
variance = colMeans(sweep(pred_matrix, 2, mean_pred, "-")^2)
mse = colMeans((pred_matrix - f0)^2)
results = data.frame(k = k_values, bias = bias, squared_bias = bias^2,
                     variance = variance, MSE = mse)
results
max_difference = max(abs(mse - (bias^2 + variance)))
max_difference
```

The variance uses divisor 200. The maximum absolute difference between
MSE and squared bias plus variance is
`r format(max_difference, scientific = TRUE, digits = 3)`, verifying the
decomposition up to numerical rounding.

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
q1_plot_data = data.frame(
  k = k_values,
  squared_bias = bias^2,
  variance = variance,
  MSE = mse
)
matplot(
  k_values, cbind(bias^2, variance, mse),
  type = "b", pch = c(1, 2, 3), lty = 1,
  col = c("darkorange", "deepskyblue3", "black"),
  xlab = "Number of neighbors k", ylab = "Squared error",
  ylim = c(0, max(bias^2, variance, mse))
)
legend(
  "topright", legend = c("Squared bias", "Variance", "MSE"),
  col = c("darkorange", "deepskyblue3", "black"),
  lty = 1, pch = c(1, 2, 3)
)
```

As $k$ increases, variance generally decreases because more observations
are averaged, while bias tends to increase because more distant observations
are included. An intermediate $k$ can therefore minimize MSE. Similarly,
stronger regularization in ridge regression and lasso reduces model
flexibility and variance but can increase bias.

## Question 2: Compare KNN with lasso

### (a)

```{r}
library(FNN)
library(glmnet)
set.seed(43252)
n = 400
p = 30
lambda = 10^seq(0, -4, length.out = 41)
results = data.frame(Setting = c("Independent", "Correlated"),
                     KNN_MSE = NA, Lasso_MSE = NA, Lambda = NA, Nonzero = NA)
for (l in 1:2) {
  Sigma = diag(p)
  if (l == 2) Sigma = 0.8 + 0.2*diag(p)
  X = matrix(rnorm((n + 1000)*p), ncol = p) %*% chol(Sigma)
  y = 0.5*X[,1] + sin(X[,2]) - 0.3*X[,3]^2 +
    rnorm(n + 1000, sd = 0.1)
  test.X = X[401:1400,]
  test.y = y[401:1400]
  X = X[1:400,]
  y = y[1:400]
  # KNN
  knn.fit = knn.reg(X, test.X, y, k = 5, algorithm = "brute")
  results$KNN_MSE[l] = mean((test.y - knn.fit$pred)^2)
  # 10-fold CV for lasso
  fold = sample(rep(1:10, each = 40))
  cv.mse = matrix(NA, 10, length(lambda))
  for (m in 1:10) {
    X.mean = colMeans(X[fold != m,])
    X.fit = sweep(X[fold != m,], 2, X.mean, "-")
    X.sd = sqrt(colMeans(X.fit^2))
    X.fit = sweep(X.fit, 2, X.sd, "/")
    X.valid = sweep(X[fold == m,], 2, X.mean, "-")
    X.valid = sweep(X.valid, 2, X.sd, "/")
    y.mean = mean(y[fold != m])
    lasso.fit = glmnet(X.fit, y[fold != m] - y.mean, alpha = 1,
                       lambda = lambda, intercept = FALSE, standardize = FALSE)
    y.hat = predict(lasso.fit, X.valid, s = lambda) + y.mean
    cv.mse[m,] = colMeans((y[fold == m] - y.hat)^2)
  }
  lambda.min = lambda[which.min(colMeans(cv.mse))]
  # Refit lasso on full training data
  X.mean = colMeans(X)
  X.fit = sweep(X, 2, X.mean, "-")
  X.sd = sqrt(colMeans(X.fit^2))
  X.fit = sweep(X.fit, 2, X.sd, "/")
  test.X.fit = sweep(test.X, 2, X.mean, "-")
  test.X.fit = sweep(test.X.fit, 2, X.sd, "/")
  y.mean = mean(y)
  lasso.fit = glmnet(X.fit, y - y.mean, alpha = 1, lambda = lambda,
                     intercept = FALSE, standardize = FALSE)
  test.pred = as.vector(predict(lasso.fit, test.X.fit, s = lambda.min)) + y.mean
  results$Lasso_MSE[l] = mean((test.y - test.pred)^2)
  results$Lambda[l] = lambda.min
  results$Nonzero[l] =
    sum(coef(lasso.fit, s = lambda.min)[-1,1] != 0)
}
results
```

### (b)

Lasso has a lower test MSE with independent predictors, whereas KNN has a
lower test MSE with correlated predictors. With independent predictors,
many irrelevant variables affect the Euclidean distance, which can make
KNN choose less useful neighbors. With correlated predictors, the variables
share more information, so the neighborhoods may better reflect the relevant
predictors. Lasso is different because it can shrink some coefficients to
zero instead of using every predictor equally.

### (c)

The true model contains nonlinear terms, $\sin(X_2)$ and $X_3^2$, while the
lasso model uses only linear terms. Therefore, lasso cannot represent the
true response function exactly. KNN does not assume a linear form, but its
performance depends on the quality of the neighbors. A single dataset in
each setting is not enough to conclude that correlation always helps KNN
or that one method is always better.

## Question 3: Observed dimension and latent dimension

### (a)

```{r}
library(FNN)
set.seed(43254)
p = 100
m_values = c(3, 30)
k_values = seq(2, 82, 8)
test_mse = matrix(NA, length(k_values), 2)
for (m in m_values) {
  A = matrix(runif(p*m, -2, 2), p, m)
  Z = matrix(rnorm(400*m), 400, m)
  X = Z %*% t(A) + matrix(rnorm(400*p), 400, p)
  y = Z[,1] + Z[,2] + rnorm(400)
  for (k in k_values) {
    knn.fit = knn.reg(X[1:200,], X[201:400,], y[1:200],
                      k = k, algorithm = "brute")
    test_mse[match(k, k_values), match(m, m_values)] =
      mean((y[201:400] - knn.fit$pred)^2)
  }
}
data.frame(k = k_values, m3 = test_mse[,1], m30 = test_mse[,2])
```

### (b)

```{r}
nsim = 50
test_mse = array(NA, c(nsim, length(k_values), 2))
for (l in 1:nsim) {
  for (m in m_values) {
    A = matrix(runif(p*m, -2, 2), p, m)
    Z = matrix(rnorm(400*m), 400, m)
    X = Z %*% t(A) + matrix(rnorm(400*p), 400, p)
    y = Z[,1] + Z[,2] + rnorm(400)
    for (k in k_values) {
      knn.fit = knn.reg(X[1:200,], X[201:400,], y[1:200],
                        k = k, algorithm = "brute")
      test_mse[l, match(k, k_values), match(m, m_values)] =
        mean((y[201:400] - knn.fit$pred)^2)
    }
  }
}
mean_mse = apply(test_mse, c(2,3), mean)
```

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
matplot(
  k_values,
  mean_mse,
  type = "b",
  pch = c(1, 2),
  lty = 1,
  col = c("deepskyblue3", "darkorange"),
  xlab = "Number of neighbors k",
  ylab = "Average test MSE"
)
legend(
  "topright",
  c("m = 3", "m = 30"),
  col = c("deepskyblue3", "darkorange"),
  lty = 1,
  pch = c(1, 2)
)
```

### (c)

When $m=3$, most of the structured variation in the 100 observed predictors
is driven by only three latent variables, so distances can better reflect
the relevant latent structure. When $m=30$, many additional latent directions
affect distances without affecting the response, making KNN neighborhoods
less informative.

The noise $\boldsymbol{\eta}_i$ also changes the observed distances. Without
this noise, $\mathbf{A}\mathbf{Z}_i$ lies in the column space of $\mathbf{A}$,
whose dimension is at most $m$. Independent noise in all 100 coordinates
means that the observed data are not confined to that space.

## Question 4: Classify handwritten digits

### (a)

```{r}
library(ElemStatLearn)
euclidean_distance = function(x0, X) {
  sqrt(rowSums(sweep(X, 2, x0, "-")^2))
}
manhattan_distance = function(x0, X) {
  rowSums(abs(sweep(X, 2, x0, "-")))
}
X = zip.train[1:1000, -1]
y = zip.train[1:1000, 1]
X_test = zip.test[1:500, -1]
y_test = zip.test[1:500, 1]
knn_classify = function(X, y, X_test, distance) {
  y_hat = integer(nrow(X_test))
  for (i in 1:nrow(X_test)) {
    d = distance(X_test[i,], X)
    neighbors = order(d, 1:nrow(X))[1:5]
    y_hat[i] =
      which.max(tabulate(y[neighbors] + 1, nbins = 10)) - 1
  }
  y_hat
}
```

### (b)

```{r}
#| echo: false
y_hat = cbind(
  Euclidean = knn_classify(X, y, X_test, euclidean_distance),
  Manhattan = knn_classify(X, y, X_test, manhattan_distance)
)
for (l in 1:2) {
  cat(colnames(y_hat)[l], "test error:", mean(y_hat[,l] != y_test), "\n")
  print(table(Predicted = factor(y_hat[,l], levels = 0:9),
              True = factor(y_test, levels = 0:9)))
}
```

Euclidean distance gives a test error of 0.122, while Manhattan distance
gives a test error of 0.138. The most common confusion is a true digit 2
being predicted as 0. Other frequent errors include 5 predicted as 0 or 3,
9 predicted as 7, and 4 predicted as 9.

### (c)

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3
#| results: hide
incorrect = which(y_hat[,1] != y_test)[1:3]
old_par = par(mfrow = c(1,3), mar = c(1,1,3,1))
for (i in incorrect) {
  image(
    zip2image(zip.test, i),
    col = gray(256:0/256),
    zlim = c(0,1),
    xlab = "",
    ylab = "",
    axes = FALSE,
    main = paste(
      "True:", y_test[i],
      "Predicted:", y_hat[i,1]
    )
  )
}
par(old_par)
```

```{r}
#| echo: false
sum(y_hat[,1] != y_hat[,2])
table(
  Euclidean_correct = y_hat[,1] == y_test,
  Manhattan_correct = y_hat[,2] == y_test
)
```

The first three Euclidean errors are 6 predicted as 0, 2 predicted as 0,
and 4 predicted as 7. These mistakes are visually understandable because
the shapes are somewhat unusual and resemble the predicted digits.

The two distance measures give different predictions on 22 images.
Euclidean is correct while Manhattan is wrong on 11 images, while the
reverse occurs on 3 images. On the remaining 8 images, both methods are
wrong but predict different digits. Therefore, two methods can have similar
overall error rates while still making mistakes on different observations.

## Question 5: Create a plot-formatting skill

### (a)

The skill is saved in `format-homework-plots/SKILL.md`. Its description
limits its use to STAT 432 homework plots and requires an explicit request.

### (b)

I asked the agent to install the skill and apply it to the Question 1 plot.
The original plot is retained in Question 1. The revised version below
uses the same plotted values and axis ranges, with a title, thicker lines,
distinct line types and point shapes, and a legend above the plotting area.

```{r}
#| label: q1-formatted
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
old_par = par(
  mar = c(4.8, 5.0, 5.0, 1.2),
  mgp = c(2.8, 0.7, 0),
  cex.axis = 1.0,
  cex.lab = 1.1,
  cex.main = 1.2,
  las = 1,
  bty = "l"
)
plot_colors = c("#D55E00", "#0072B2", "#333333")
plot_types = c(2, 3, 1)
plot_points = c(16, 17, 15)
matplot(
  q1_plot_data$k,
  as.matrix(q1_plot_data[c("squared_bias", "variance", "MSE")]),
  type = "b",
  lwd = 2,
  pch = plot_points,
  lty = plot_types,
  cex = 0.85,
  col = plot_colors,
  xlab = "Number of neighbors k",
  ylab = "Squared error",
  xaxt = "n",
  ylim = c(0, max(q1_plot_data[c("squared_bias", "variance", "MSE")]))
)
axis(1, at = q1_plot_data$k[seq(1, nrow(q1_plot_data), by = 2)])
title(main = "KNN bias-variance trade-off", line = 2.5)
legend(
  "top",
  inset = c(0, -0.17),
  legend = c("Squared bias", "Variance", "MSE"),
  col = plot_colors,
  lty = plot_types,
  pch = plot_points,
  lwd = 2,
  pt.cex = 0.85,
  cex = 0.9,
  x.intersp = 1.4,
  horiz = TRUE,
  bty = "n",
  xpd = NA
)
par(old_par)
```

### (c)

I chose these rules to improve readability while preserving the original
analysis. I prefer the revised plot because it is easier to read and makes
the differences between squared bias, variance, and MSE clearer. The
formatting changes did not alter any data, calculations, or results, so I
decided to keep the skill.

```{=latex}
\newpage
```

The final `SKILL.md` content is:

```markdown
---
name: format-homework-plots
description: Use only when the user explicitly requests this skill to improve the formatting of STAT 432 homework plots in R or Python.
---

# STAT 432 Plot Formatting

- Preserve the original data, calculations, and statistical meaning.
- Make important trends and differences visually clear without exaggerating them.
- Use sufficient margins so titles, axis labels, tick labels, and legends are not clipped.
- Use a short and informative title.
- Use clear axis labels and readable tick labels.
- Use high-contrast, distinguishable colors for multiple curves, and avoid colors that are too similar.
- Use different point shapes or line types when multiple curves appear in the same plot.
- Use thicker lines and appropriately sized points so important details are easy to see.
- Use readable font sizes for titles, axes, and legends.
- Keep the figure size balanced and avoid unnecessary visual clutter.
- Apply these rules to existing R or Python plotting code without changing the underlying analysis.
```
